---
title: "Complicate Steady States Analysis"
author: "Christope Dutang"
date: "May 2016"
output: rmarkdown::html_vignette
bibliography: markovchainBiblio.bib
vignette: >
  %\VignetteIndexEntry{complicate_steady_states_analysis} 
  %\VignetteEngine{knitr::rmarkdown}  
  %VignetteEncoding{UTF-8}
---

# Example 1 

## Transition matrix
Let us consider the Markov chain with the following transition matrix
$$
M = \left(\begin{matrix}
0 & 1/2 & 1/2 & 0 & 0 & 0 \\
0 & 0 & 1-p & 0 & 0 & p \\
0 & 1-q & 0 & q & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 \\
\end{matrix}\right)
$$
with $1>p>0$ close to zero, e.g. $p=1\%$ and $1>q>0$, e.g. $q=0.2$.

## R implementation 
From the diagram or $M$, it is easy that there are absorbing states (6) and recurrent states (4,5).
```{r, message=FALSE}
library(markovchain)
M<-matrix(c(0,0.5,0.50,0.0,0,0.00,0,0.0,0.99,0.0,0,0.01,0,0.8,0.00,
                0.2,0,0.00,0,0.0,0.00,0.0,1,0.00,
0,0.0,0.00,1.0,0,0.00,0,0.0,0.00,0.0,0,1.00), nrow=6, byrow=TRUE)
colnames(M) <- rownames(M) <- paste0("s",1:6)
chain1 <- as(M,"markovchain")
plot(chain1)
```

## Invariant measure
We need to solve the following system in order to get the invariant measure
$$
\mu M = \mu
\Leftrightarrow
\left\{\begin{array}{l}
0 = \mu_1 \\
0.5\mu_1 + (1-q)\mu_3 = \mu_2 \\
0.5\mu_1 + (1-p)\mu_2 = \mu_3 \\
q\mu_3+\mu_5= \mu_4 \\
\mu_4=\mu_5 \\
p\mu_2 + \mu_6=\mu_6
\end{array}\right.
\Leftrightarrow
\left\{\begin{array}{l}
\mu_1 = 0  \\
 (1-q)\mu_3 = \mu_2 \\
 (1-p)\mu_2 = \mu_3 \\
q\mu_3+\mu_4= \mu_4 \\
\mu_5=\mu_4 \\
p\mu_2 =0
\end{array}\right.
\Leftrightarrow
\left\{\begin{array}{l}
\mu_1 = 0  \\
 \mu_2=0 \\
 \mu_3=0 \\
\mu_5=\mu_4 \\
\end{array}\right.
$$
Using the probability condition $\sum_i \mu_i=1$, we get $2\mu_4+\mu_6=1$.
So there exists an infinite number of invariant measures of the form (as long as $p>0$)
$$
\mu = (0, 0, 0, \nu, \nu, 1-2\nu), \nu\in[0,1/2].
$$
Note that there are independent of the value of $p$ and $q$.
Indeed, we have
```{r}
mu <- c(0, 0, 0, 1/4, 1/4, 1/2)
mu %*% M 
mu <- c(0, 0, 0, 0, 0, 1)
mu %*% M 
mu <- c(0, 0, 0, 1/2, 1/2, 0)
mu %*% M 
steadyStates(chain1)
```


## Classification of states
Let $T^{x\rightarrow x}$ is the number of periods to go back to state $x$ knowing that the chain starts in $x$.  

* A state $x$ is recurrent if $P(T^{x\rightarrow x}<+\infty)=1$ (equivalently $P(T^{x\rightarrow x}=+\infty)=0$).  
    * A state $x$ is null recurrent if in addition $E(T^{x\rightarrow x})=+\infty$.
    * A state $x$ is positive recurrent if in addition $E(T^{x\rightarrow x})<+\infty$.
    * A state $x$ is absorbing if in addition  $P(T^{x\rightarrow x}=1)=1$.  
* A state $x$ is transient if $P(T^{x\rightarrow x}<+\infty)<1$ (equivalently $P(T^{x\rightarrow x}=+\infty)>0$).  


On the example above, using the transition matrix, we have
$$
\begin{array}{l}
\forall k>0, P(T^{s1\rightarrow s1}=k) = 0 \Rightarrow P(T^{s1\rightarrow s1}=+\infty)=1, \\
P(T^{s4\rightarrow s4}=2) = 1  \Rightarrow P(T^{s4\rightarrow s4}=+\infty)=0,\\
P(T^{s5\rightarrow s5}=2) = 1  \Rightarrow P(T^{s5\rightarrow s5}=+\infty)=0,\\
P(T^{s6\rightarrow s6}=1) = 1  \Rightarrow P(T^{s6\rightarrow s6}=+\infty)=0.
\end{array}
$$
So the states $s4, s5, s6$ are recurrent (since $E(T^{s4\rightarrow s4})=2=E(T^{s5\rightarrow s5})$
and $E(T^{s6\rightarrow s6})=1$, they are positive recurrent), whereas the state $s1$ is transient.
Furthermore, for the last two states, we have
$$
\begin{array}{l}
P(T^{s2\rightarrow s2}=2k) = 0.8^k(1-p)^k, p>0 \Rightarrow P(T^{s2\rightarrow s2}=+\infty)=0, \\
P(T^{s3\rightarrow s3}=2k) = (1-p)^k0.8^k, p>0 \Rightarrow P(T^{s3\rightarrow s3}=+\infty)=0. \\
\end{array}
$$
So the states $s2, s3$ are recurrent (since $E(T^{s2\rightarrow s2})=E(T^{s3\rightarrow s3})=1/(1-0.8(1-p))<+\infty$, they are positive recurrent).


## Analysis of the eigenvalues and eigenvectors
Let us compute the characteristic polynom of the matrix $M$
\begin{eqnarray*}
\chi_M(x) &=&
\left|\begin{matrix}
-x & 1/2 & 1/2 & 0 & 0 & 0 \\
0 & -x & 1-p & 0 & 0 & p \\
0 & 1-q & -x & q & 0 & 0 \\
0 & 0 & 0 & -x & 1 & 0 \\
0 & 0 & 0 & 1 & -x & 0 \\
0 & 0 & 0 & 0 & 0 & 1-x \\
\end{matrix}\right|
= 
\left|\begin{matrix}
-x & 1/2 & 1/2 & 0 & 0  \\
0 & -x & 1-p & 0 & 0  \\
0 & 1-q & -x & q & 0  \\
0 & 0 & 0 & -x & 1  \\
0 & 0 & 0 & 1 & -x  \\
\end{matrix}\right|
(1-x)
\\
&=& 
-(1-x)
\left|\begin{matrix}
-x & 1/2 & 1/2  & 0  \\
0 & -x & 1-p  & 0  \\
0 & 1-q & -x  & 0  \\
0 & 0 & 0  & 1  \\
\end{matrix}\right|
-x(1-x)
\left|\begin{matrix}
-x & 1/2 & 1/2 & 0   \\
0 & -x & 1-p & 0   \\
0 & 1-q & -x & q   \\
0 & 0 & 0 & -x   \\
\end{matrix}\right|
\\
&=& 
-(1-x)
\left|\begin{matrix}
-x & 1/2 & 1/2    \\
0 & -x & 1-p    \\
0 & 1-q & -x    \\
\end{matrix}\right|
+x^2(1-x)
\left|\begin{matrix}
-x & 1/2 & 1/2    \\
0 & -x & 1-p    \\
0 & 1-q & -x    \\
\end{matrix}\right|
\\
&=& 
+x(1-x)
\left|\begin{matrix}
 -x & 1-p    \\
 1-q & -x    \\
\end{matrix}\right|
-x^3(1-x)
\left|\begin{matrix}
 -x & 1-p    \\
 1-q & -x    \\
\end{matrix}\right|
\\
&=& 
[x(1-x)
-x^3(1-x)
](x^2-(1-p)(1-q))
=
x(1-x)^2(1+x)(x-(1-p)(1-q))(x+(1-p)(1-q))
\end{eqnarray*}
Therefore the eigenvalues are $\pm 1, 0, \pm(1-p)(1-q)$ with 1 of multiplicity 2.
Hence the power of $M$ is 
$$
M^n  = P \left|\begin{matrix}
1^n & 0 & 0 & 0 & 0 & 0 \\
0 & 1^n & 0 & 0 & 0 & 0 \\
0 & 0 & (-1)^n & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & (1-p)^n(1-q)^n & 0 \\
0 & 0 & 0 & 0 & 0 & (1-p)^n(1-q)^n \\
\end{matrix}\right| P^{-1}
$$
with an appropriate matrix $P$ changing base in $\mathbb R^6$.

## Numerical consideration

The classification of states is given by dedicated functions
```{r}
communicatingClasses(chain1)
recurrentClasses(chain1)
absorbingStates(chain1)
transientStates(chain1)
```

Let us compute power of $M$. Note that even powers of $M$ yield to a matrix
different than odd powers since $P(T^{s4\rightarrow s4}=2)=P(T^{s5\rightarrow s5}=2)=1$.
```{r}
library(expm)
M %^% 1000
M %^% 1001
```


Invariant measures are the eigenvectors associated to the eigen value 1 
for the transposed matrix $M^T$ such that $\mu_i\geq 0$. If in addition $\sum_i \mu_i=1$,
then it is a probability measure.

Let us compute the eigenvalue. We don't retrieve the theoretical values $\pm 1, 0, \pm(1-p)(1-q)$.
```{r}
c(1, -1, 0, .99*.8, -.99*.8)
eigen(t(M))
```
Keeping those equalling 1, we obtain
```{r}
eigen(t(M))$vectors[,eigen(t(M))$values >= 1]
```
So numerically, only one (probability) measure is obtained $(0,0,0,0,0,1)$. That's rather logical.

One can get the "boundary" measures by looking to the subset of recurrent states, extracting positive eigenvectors associated to the eigenvalue 1 and normalize ti. Indeed
```{r}
Msub <- M[rownames(M) %in% unlist(recurrentClasses(chain1)), colnames(M) %in% unlist(recurrentClasses(chain1))]
eigen(Msub)$vectors[,eigen(Msub)$values >= 1]
t(eigen(Msub)$vectors[,eigen(Msub)$values >= 1]) / colSums(eigen(Msub)$vectors[,eigen(Msub)$values >= 1])
```

The `steadyStates` function in [@pkg:markovchain] implements the abovementioned algorithm when negative values are found in the eigenvectors of the full matrix corresponding to unitary eigenvalues.

# References