---
title: "Higher order Markov chains"
author: "Deepak Yadav, Tae Seung Kang, Giorgio Alfredo Spedicato"
format: html
bibliography:
  - markovchainBiblio.bib
vignette: >
  %\VignetteIndexEntry{Higher order Markov chains}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: global_options
#| include: false
knitr::opts_chunk$set(fig.width=8.5, fig.height=6, out.width = "70%")
set.seed(123)
# ANSI colour codes (U+001B) break the LaTeX build, so keep the output plain
options(crayon.enabled = FALSE, cli.num_colors = 1, cli.hyperlink = FALSE)
Sys.setenv(NO_COLOR = "1")
```

# Higher Order Markov Chains

Continuous time Markov chains are discussed in the CTMC vignette which is a part of the package.

An experimental `fitHigherOrder` function has been written in order to fit a higher order Markov chain ([@ching2013higher, @ching2008higher]). 
`fitHigherOrder` takes three inputs

  1. sequence: a categorical data sequence.
  2. order: order of Markov chain to fit with default value 2.
  3. method: how the weights are estimated, `"lsq"` (default) or `"mle"`, see below.

The output will be a `list` which consists of

  1. lambda: model parameter(s).
  2. Q: a list of transition matrices. $Q_i$ is the $ith$ step transition matrix stored column-wise.
  3. X: frequency probability vector of the given sequence.

The model is a mixture of the empirical lag-$i$ transition matrices $Q_i$ with weights $\lambda_i\ge0$ summing to one, and the matrices $Q_i$ are the same for both methods. With the default `method = "lsq"` the weights solve a quadratic programming problem, the minimization of the squared distance between the stationary distribution and its image under the mixture, which is solved using `solnp` function of the Rsolnp package [@pkg:Rsolnp]. With `method = "mle"` the weights maximize the log-likelihood of the observations; for fixed $Q_i$ this is a concave problem, so its maximum is global, and it is solved by the EM algorithm for mixture weights, without the Rsolnp package.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      comment = "#>")
```

```{r}
#| label: setup_2
#| include: false
#| message: false
#| echo: false
require(markovchain)
```


```{r}
#| label: higherOrder
#| message: false
#| warning: false
if (requireNamespace("Rsolnp", quietly = TRUE)) {
  data(rain)
  rain_small <- rain$rain[1:150]
  fitHigherOrder(rain_small, 2)
}
```

# Comparing models of different orders

`higherOrderLogLik` evaluates the log-likelihood of a sequence under the model returned by `fitHigherOrder`, and derives the deviance ($-2$ times the log-likelihood), the AIC and the BIC, so that models of different orders can be compared. The probability of moving to state $x_t$ is the mixture $\sum_{i=1}^{k}\lambda_i Q_i[x_t, x_{t-i}]$, the form of the mixture transition distribution model of [@raftery1985model], and the log-likelihood is the sum of the logarithms of these probabilities over the observations that a model of order $k$ can predict, that is from observation $k+1$ onwards. To compare orders on exactly the same observations the argument `start` must be set to one plus the largest order compared.

Two caveats apply. First, with the default `method = "lsq"` the weights $\lambda$ are chosen by least squares on the stationary distribution, not by maximum likelihood, so the value returned is the log-likelihood *of the fitted model* and not the maximum attainable one. It can therefore be lower for a higher order than for a lower one, which cannot happen for maximum-likelihood fits of nested models. With `method = "mle"` the weights maximize this log-likelihood for the observations that a model of that order can predict, and in the examples below the log-likelihood then never decreases with the order. Second, the number of parameters used for the criteria is $k\,r(r-1)+(k-1)$, with $r$ the number of states.

The example compares orders one to three on the Alofi Island daily rainfall and on the preproglucacon DNA sequence, both analysed by [@averyHenderson], with both estimation methods. Maximum likelihood weights raise the log-likelihood of the higher orders, by up to about 11 units for the rainfall and 28 for the DNA sequence, whereas the least squares weights can leave it below that of the first-order model. Even so, the gain is too small to pay for the additional parameters, and both criteria select the first-order model for both sequences with both methods. The values are computed by the package; they are not claimed to reproduce those of the original paper.

```{r}
#| label: compareOrders
#| message: false
#| warning: false
if (requireNamespace("Rsolnp", quietly = TRUE)) {
  compareOrders <- function(sequence, orders = 1:3) {
    byMethod <- lapply(c("lsq", "mle"), function(method) {
      fits <- lapply(orders, function(k) fitHigherOrder(sequence, k, method = method))
      sapply(fits, function(f)
        unlist(higherOrderLogLik(sequence, f, start = max(orders) + 1)[c("logLik", "BIC")]))
    })
    out <- rbind(byMethod[[1]], byMethod[[2]])
    rownames(out) <- c("logLik (lsq)", "BIC (lsq)", "logLik (mle)", "BIC (mle)")
    colnames(out) <- paste("order", orders)
    round(out, 1)
  }
  data(rain)
  print(compareOrders(rain$rain))
  data(preproglucacon)
  print(compareOrders(preproglucacon$preproglucacon))
}
```

## Reproducing a published comparison

[@berchtold2002mixture] compare, by log-likelihood and BIC, independence, Markov chains of order one to three and mixture transition distribution (MTD) models for the hourly wind direction at Koeberg (South Africa; 744 observations recoded into four directions, originally from [@macdonald1997hidden]) and for a daily series of epileptic seizures (204 observations). The authors kindly provided the two series, which are distributed with the package in `inst/extdata`. Their convention is to condition every model on the first 14 observations, so that all models are evaluated on the same $n-14$ observations, which in `higherOrderLogLik` corresponds to `start = 15`; the BIC uses $n-14$ as sample size and counts only the parameters that are not forced to zero.

`fitHigherOrder` does not estimate the MTD model of that paper: it fits a different transition matrix for each lag, with weights chosen by least squares or, with `method = "mle"`, by maximum likelihood given those matrices, whereas the MTD model uses a single matrix $Q$ for all lags and estimates it together with the weights by maximum likelihood. The comparison below therefore evaluates, with `higherOrderLogLik`, the first-order chain estimated on the observations entering the likelihood and the MTD(2) model with the weights and the matrix $Q$ printed in the paper (whose rows are the departure states, hence the transposition).

```{r}
#| label: berchtoldRaftery
koeberg <- as.character(read.csv(system.file("extdata", "koeberg_wind.csv",
                                             package = "markovchain"))$state)
start <- 15
n_eff <- length(koeberg) - (start - 1)

# first-order Markov chain, estimated on the transitions that enter the likelihood
Q1 <- seq2matHigh(koeberg[(start - 1):length(koeberg)], 1)
mc1 <- higherOrderLogLik(koeberg, list(lambda = 1, Q = list(Q1)), start = start)

# MTD(2) with lambda and Q as printed in Section 1.3 of the paper
Qpaper <- matrix(c(0.8301, 0.0689, 0.0077, 0.0933,
                   0.0369, 0.9012, 0.0619, 0.0000,
                   0.0155, 0.1553, 0.8070, 0.0222,
                   0.0779, 0.0000, 0.0528, 0.8693), 4, 4, byrow = TRUE)
Q <- t(Qpaper)
dimnames(Q) <- list(as.character(1:4), as.character(1:4))
mtd2 <- higherOrderLogLik(koeberg, list(lambda = c(0.7569, 0.2431), Q = list(Q, Q)),
                          start = start)

# parameters not forced to zero: 11 for the chain (one empty transition),
# 4 * 3 - 2 + (2 - 1) = 11 for the MTD(2) (two structural zeros in Q)
comparison <- data.frame(
  model = c("Markov chain, order 1", "MTD, order 2"),
  logLik = c(mc1$logLik, mtd2$logLik),
  BIC = c(-2 * mc1$logLik + 11 * log(n_eff), -2 * mtd2$logLik + 11 * log(n_eff)),
  logLik_published = c(-413.3, -393.4),
  BIC_published = c(899.1, 859.3))
print(comparison, digits = 4, row.names = FALSE)
```

The values coincide with Table 2 of the paper up to its rounding to one decimal, and the BIC prefers the MTD(2) model to the first-order chain, as in the paper. The same agreement is obtained for the Markov chains of order two and three and for the seizure series (Table 3); these checks are part of the unit tests of the package.

# Higher Order Multivariate Markov Chains

## Introduction

HOMMC model is used for modeling behaviour of multiple categorical sequences generated by similar sources. The main reference is [@ching2008higher]. Assume that there are s categorical sequences and each has possible states in M. In nth order MMC the state probability distribution of the jth sequence at time $t = r + 1$ depend on the state probability distribution of all the sequences (including itself) at times $t = r, r - 1, ..., r - n + 1$.

\[
x_{r+1}^{(j)} = \sum_{k=1}^{s}\sum_{h=1}^{n}\lambda_{jk}^{(h)}P_{h}^{(jk)}x_{r-h+1}^{(k)}, j = 1, 2, ..., s, r = n-1, n, ...
\]

with initial distribution $x_{0}^{(k)}, x_{1}^{(k)}, ... , x_{n-1}^{(k)} (k = 1, 2, ... , s)$. Here

\[
\lambda _{jk}^{(h)} \geq 0, 1\leq j, k\leq s, 1\leq h\leq  n \enspace and \enspace \sum_{k=1}^{s}\sum_{h=1}^{n} \lambda_{jk}^{(h)} = 1, j = 1, 2, 3, ... , s.
\]

Now we will see the simpler representation of the model which will help us understand the result of `fitHighOrderMultivarMC` method.

\vspace{5mm}

Let $X_{r}^{(j)} = ((x_{r}^{(j)})^{T}, (x_{r-1}^{(j)})^{T}, ..., (x_{r-n+1}^{(j)})^{T})^{T} for \enspace j = 1, 2, 3, ... , s.$ Then

\vspace{5mm}

\[
\begin{pmatrix}
X_{r+1}^{(1)}\\ 
X_{r+1}^{(2)}\\ 
.\\ 
.\\ 
.\\ 
X_{r+1}^{(s)}
\end{pmatrix} = \begin{pmatrix}
 B^{11}&  B^{12}&  .&  .&  B^{1s}& \\ 
 B^{21}&  B^{22}&  .&  .&  B^{2s}& \\ 
 .&  .&  .&  .&  .& \\ 
 .&  .&  .&  .&  .& \\ 
 .&  .&  .&  .&  .& \\ 
 B^{s1}&  B^{s2}&  .&  .&  B^{ss}& \\ 
\end{pmatrix} \begin{pmatrix}
X_{r}^{(1)}\\ 
X_{r}^{(2)}\\ 
.\\ 
.\\ 
.\\ 
X_{r}^{(s)}
\end{pmatrix} \textrm{where}
\]

\[B^{ii} = \begin{pmatrix}
 \lambda _{ii}^{(1)}P_{1}^{(ii)}&  \lambda _{ii}^{(2)}P_{2}^{(ii)}&  .&  .&  \lambda _{ii}^{(n)}P_{n}^{(ii)}& \\ 
 I&  0&  .&  .&  0& \\ 
 0&  I&  .&  .&  0& \\ 
 .&  .&  .&  .&  .& \\ 
 .&  .&  .&  .&  .& \\ 
 0&  .&  .&  I&  0& 
\end{pmatrix}_{mn*mn} \textrm{and}
\]

\vspace{5mm}

\[
B^{ij} = \begin{pmatrix}
 \lambda _{ij}^{(1)}P_{1}^{(ij)}&  \lambda _{ij}^{(2)}P_{2}^{(ij)}&  .&  .&  \lambda _{ij}^{(n)}P_{n}^{(ij)}& \\ 
 0&  0&  .&  .&  0& \\ 
 0&  0&  .&  .&  0& \\ 
 .&  .&  .&  .&  .& \\ 
 .&  .&  .&  .&  .& \\ 
 0&  .&  .&  0&  0& 
\end{pmatrix}_{mn*mn} \textrm{when } i\neq j.
\]

\vspace{5mm}

## Representation of parameters in the code

$P_{h}^{(ij)}$ is represented as $Ph(i,j)$ and $\lambda _{ij}^{(h)}$ as Lambdah(i,j). 
For example: $P_{2}^{(13)}$ as $P2(1,3)$ and $\lambda _{45}^{(3)}$ as Lambda3(4,5).

## Definition of HOMMC class

```{r}
#| label: hommcObject
showClass("hommc")
```

Any element of `hommc` class is comprised by following slots:

  1. states: a character vector, listing the states for which transition probabilities are defined.
  2. byrow: a logical element, indicating whether transition probabilities are shown by row or by column.
  3. order: order of Multivariate Markov chain.
  4. P: an array of all transition matrices.
  5. Lambda: a vector to store the weightage of each transition matrix.
  6. name: optional character element to name the HOMMC

## How to create an object of class HOMMC

```{r}
#| label: hommcCreate
states <- c('a', 'b')
P <- array(dim = c(2, 2, 4), dimnames = list(states, states))
P[ , , 1] <- matrix(c(1/3, 2/3, 1, 0), byrow = FALSE, nrow = 2, ncol = 2)

P[ , , 2] <- matrix(c(0, 1, 1, 0), byrow = FALSE, nrow = 2, ncol = 2)

P[ , , 3] <- matrix(c(2/3, 1/3, 0, 1), byrow = FALSE, nrow = 2, ncol = 2)

P[ , , 4] <- matrix(c(1/2, 1/2, 1/2, 1/2), byrow = FALSE, nrow = 2, ncol = 2)

Lambda <- c(.8, .2, .3, .7)

hob <- new("hommc", order = 1, Lambda = Lambda, P = P, states = states, 
           byrow = FALSE, name = "FOMMC")
hob
```

## Fit HOMMC

`fitHighOrderMultivarMC` method is available to fit HOMMC. Below are the 3 parameters of this method.

  1. seqMat: a character matrix or a data frame, each column represents a categorical sequence.
  2. order: order of Multivariate Markov chain. Default is 2.
  3. Norm: Norm to be used. Default is 2.

# A Marketing Example

We tried to replicate the example found in [@ching2008higher] for an application of HOMMC. A soft-drink company in Hong Kong is facing an in-house problem of production planning and inventory control. A pressing issue is the storage space of its central warehouse, which often finds itself in the state of overflow or near capacity. The company is thus in urgent needs to study the interplay between the storage space requirement and the overall growing sales demand. The product can be classified into six possible states (1, 2, 3, 4, 5, 6) according to their sales volumes. All products are labeled as 1 = no sales volume, 2 = very slow-moving (very low sales volume), 3 = slow-moving, 4 = standard, 5 = fast-moving or 6 = very fast-moving (very high sales volume). Such labels are useful from both marketing and production planning points of view. The data is cointaind in `sales` object.

```{r}
#| label: hommsales
data(sales)
head(sales)
```

The company would also like to predict sales demand for an important customer in order to minimize its inventory build-up. More importantly, the company can understand the sales pattern of this customer and then develop a marketing strategy to deal with this customer. Customer's sales demand sequences of five important products of the company for a year. We expect sales demand sequences generated by the same customer to be correlated to each other. Therefore by exploring these relationships, one can obtain a better higher-order multivariate Markov model for such demand sequences, hence obtain better prediction rules.

In [@ching2008higher] application, they choose the order arbitrarily to be eight, i.e., n = 8. We first estimate all the transition probability matrices $P_{h}^{ij}$ and we also have the estimates of the stationary probability distributions of the five products:.

$\widehat{\boldsymbol{x}}^{(1)} = \begin{pmatrix}
0.0818&  0.4052&  0.0483&  0.0335& 0.0037& 0.4275 
\end{pmatrix}^{\boldsymbol{T}}$

$\widehat{\boldsymbol{x}}^{(2)} = \begin{pmatrix}
0.3680&  0.1970&  0.0335&  0.0000& 0.0037& 0.3978 
\end{pmatrix}^{\boldsymbol{T}}$

$\widehat{\boldsymbol{x}}^{(3)} = \begin{pmatrix}
0.1450& 0.2045& 0.0186& 0.0000& 0.0037& 0.6283 
\end{pmatrix}^{\boldsymbol{T}}$

$\widehat{\boldsymbol{x}}^{(4)} = \begin{pmatrix}
0.0000& 0.3569& 0.1338& 0.1896& 0.0632& 0.2565
\end{pmatrix}^{\boldsymbol{T}}$

$\widehat{\boldsymbol{x}}^{(5)} = \begin{pmatrix}
0.0000& 0.3569& 0.1227& 0.2268& 0.0520& 0.2416
\end{pmatrix}^{\boldsymbol{T}}$

By solving the corresponding linear programming problems, we obtain the following higher-order multivariate Markov chain model:

\vspace{3mm}

$\boldsymbol{x}_{r+1}^{(1)} = \boldsymbol{P}_{1}^{(12)}\boldsymbol{x}_{r}^{(2)}$

$\boldsymbol{x}_{r+1}^{(2)} = 0.6364\boldsymbol{P}_{1}^{(22)}\boldsymbol{x}_{r}^{(2)} + 
0.3636\boldsymbol{P}_{3}^{(22)}\boldsymbol{x}_{r}^{(2)}$

$\boldsymbol{x}_{r+1}^{(3)} = \boldsymbol{P}_{1}^{(35)}\boldsymbol{x}_{r}^{(5)}$

$\boldsymbol{x}_{r+1}^{(4)} = 0.2994\boldsymbol{P}_{8}^{(42)}\boldsymbol{x}_{r}^{(2)} + 
0.4324\boldsymbol{P}_{1}^{(45)}\boldsymbol{x}_{r}^{(5)} + 0.2681\boldsymbol{P}_{2}^{(45)}\boldsymbol{x}_{r}^{(5)}$

$\boldsymbol{x}_{r+1}^{(5)} = 0.2718\boldsymbol{P}_{8}^{(52)}\boldsymbol{x}_{r}^{(2)} + 
0.6738\boldsymbol{P}_{1}^{(54)}\boldsymbol{x}_{r}^{(4)} + 0.0544\boldsymbol{P}_{2}^{(55)}\boldsymbol{x}_{r}^{(5)}$

\vspace{3mm}

According to the constructed 8th order multivariate Markov model, Products A and B are closely related. In particular, the sales demand of Product A depends strongly on Product B. The main reason is that the chemical nature of Products A and B is the same, but they have different packaging for marketing purposes. Moreover, Products B, C, D and E are closely related. Similarly, products C and E have the same product flavor, but different packaging. In this model, it is interesting to note that both Product D and E quite depend on Product B at order of 8, this relationship is hardly to be obtained in conventional Markov model owing to huge amount of parameters. The results show that higher-order multivariate Markov model is quite significant to analyze the relationship of sales demand.

```{r}
#| label: hommcFit
#| warning: false
#| message: false
#| eval: false


# fit 8th order multivariate markov chain
if (requireNamespace("Rsolnp", quietly = TRUE)) {
object <- fitHighOrderMultivarMC(sales, order = 8, Norm = 2)
}
```

We choose to show only results shown in the paper. We see that $\lambda$ values are quite close, but not equal, to those shown in the original paper.

```{r}
#| label: result
#| echo: false
#| eval: false


if (requireNamespace("Rsolnp", quietly = TRUE)) {
i <- c(1, 2, 2, 3, 4, 4, 4, 5, 5, 5)
j <- c(2, 2, 2, 5, 2, 5, 5, 2, 4, 5)
k <- c(1, 1, 3, 1, 8, 1, 2, 8, 1, 2)

if(object@byrow == TRUE) {
    direction <- "(by rows)" 
} else {
    direction <- "(by cols)"
}

cat("Order of multivariate markov chain =", object@order, "\n")
cat("states =", object@states, "\n")

cat("\n")
cat("List of Lambda's and the corresponding transition matrix", direction,":\n")

for(p in 1:10) {
    t <- 8*5*(i[p]-1) + (j[p]-1)*8
    cat("Lambda", k[p], "(", i[p], ",", j[p], ") : ", object@Lambda[t+k[p]],"\n", sep = "")
    cat("P", k[p], "(", i[p], ",", j[p], ") : \n", sep = "")
    print(object@P[, , t+k[p]])
    cat("\n")
}
} else {
  print("package Rsolnp unavailable")
}
```

## Acknowledgments

We are grateful to Professors Adrian E. Raftery and André Berchtold for kindly sharing the Koeberg wind-direction and epileptic-seizure series of [@berchtold2002mixture], which made it possible to check `higherOrderLogLik` against the published results.

## References

::: {#refs}
:::
