---
title: "Published literature benchmarks for Markov-chain tests"
author: "Giorgio Alfredo Spedicato"
format: html
bibliography:
  - statistical_analysis_refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Published literature benchmarks for Markov-chain tests}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
library(markovchain)
```

## A published non-rejection example for the Markov property: Burkett, Scherer and Todd (2019)

Burkett, Scherer and Todd study recurrent and persistent states in financial markets and use chi-squared tests to investigate the first-order Markov property [@burkett2019recurrent]. Their published examples are useful benchmarks because the observed counts used in the tests are reported explicitly.

For one of their tests, the transition of interest is $1\to1$. Observations are grouped according to the state immediately preceding the origin state. The published counts are:

```{r}
#| label: burkett-markov-counts

burkett <- matrix(
  c(127, 59,
     11,  7,
     36, 20,
      9,  4),
  nrow = 4,
  byrow = TRUE,
  dimnames = list(
    c("previous state 1", "previous state 2",
      "previous state 4", "previous state 5"),
    c("transition to state 1", "transition elsewhere")
  )
)

burkett
```

The Pearson test on these published counts does not reject the null hypothesis that the probability of the transition $1\to1$ is independent of the preceding state. The reported p-value is above 0.05 (the paper reports a non-significant result for this example). Thus this provides an independent, published example in which a Markov-property hypothesis is **not rejected**.

```{r}
#| label: burkett-markov-test

burkett.test <- stats::chisq.test(burkett, correct = FALSE)
burkett.test
```

This example is deliberately presented as a literature benchmark for the contingency-table test underlying Markov-property inference. It should not be interpreted as a direct call to `verifyMarkovProperty()` on a sequence: the paper performs a cell-wise conditional test, whereas `verifyMarkovProperty()` evaluates the first-order Markov property jointly over the observed triple-transition table.

## Published empirical-versus-theoretical transition matrix: Kullback, Kupperman and Ku (1962)

Kullback, Kupperman and Ku [@kullback1962tests] develop likelihood-ratio and information-theoretic tests for finite Markov chains, including the hypothesis that an empirical transition matrix is compatible with a **specified theoretical transition matrix**. Their paper contains a worked example that is reproduced in the `markovchain` documentation and provides exactly the type of benchmark needed for `verifyEmpiricalToTheoretical()`.

The published sequence is:

```{r}
#| label: kullback-sequence

kullback.sequence <- c(
  0,1,2,2,1,0,0,0,0,0,0,1,2,2,2,1,0,0,1,0,0,0,0,0,0,1,1,
  2,0,0,2,1,1,0,0,0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,1,0,0,0,0,2,1,0,
  0,2,1,0,0,0,0,0,0,1,1,1,2,2,0,0,2,1,1,1,1,2,1,1,1,1,1,1,1,1,1,0,2,
  0,1,1,0,0,0,0,0,0,0,0,2,2,2,0,0,0,1,1,1,1,1,2,1,2,0,0,0,1,1
)

kullback.counts <- createSequenceMatrix(kullback.sequence)
kullback.counts
```

The theoretical transition matrix specified in the published example is

```{r}
#| label: kullback-theoretical-matrix

kullback.P <- matrix(
  c(5/8, 1/4, 1/8,
    1/4, 1/2, 1/4,
    1/4, 3/8, 3/8),
  byrow = TRUE,
  nrow = 3,
  dimnames = list(as.character(0:2), as.character(0:2))
)

theoretical <- new(
  "markovchain",
  states = as.character(0:2),
  transitionMatrix = kullback.P
)

theoretical
```

The published likelihood-ratio calculation gives a chi-squared statistic of approximately $6.55$ on 6 degrees of freedom, with p-value approximately **0.3643**. Hence the null hypothesis that the empirical transition matrix is compatible with the specified theoretical matrix is **not rejected** at the 5% level [@kullback1962tests].

The same calculation can be reproduced with the package:

```{r}
#| label: kullback-empirical-theoretical

result <- verifyEmpiricalToTheoretical(
  data = kullback.sequence,
  object = theoretical,
  method = "G",
  verbose = TRUE
)

result
```

The package uses the likelihood-ratio statistic described by Kullback et al. for this test. The published result and the package result therefore provide a particularly useful regression benchmark: **the empirical transition matrix is not rejected as compatible with the specified theoretical Markov chain**.

For comparison, using the Pearson version gives the same inferential conclusion:

```{r}
#| label: kullback-empirical-theoretical-pearson

verifyEmpiricalToTheoretical(
  data = kullback.sequence,
  object = theoretical,
  method = "Pearson",
  verbose = TRUE
)
```

This example is preferable to a simulated example when demonstrating `verifyEmpiricalToTheoretical()`: both the empirical sequence and the theoretical transition matrix are taken from a published worked example, and the null hypothesis is explicitly **not rejected**.

## Why these examples are useful together

The two examples provide complementary published benchmarks for the package's statistical-inference functionality:

- **Burkett et al. (2019):** an independently published, real-data example in which a Markov-property hypothesis is not rejected.
- **Kullback et al. (1962):** a published empirical-versus-theoretical transition-matrix test with $p\approx0.364$, so the specified Markov chain is not rejected.
- **Tran and Carmichael (2012):** a real contractor-payment application in the main statistical-analysis vignette in which time-homogeneity is rejected.

Together they avoid presenting only rejection examples and demonstrate the important inferential principle that a large p-value means **failure to reject the null hypothesis**, not proof that the Markov model is true.
