---
title: "Crash Introduction to markovchain R package"
author: "Giorgio Alfredo Spedicato, Ph.D C.Stat ACAS"
date: "19th April 2015"
output:
  beamer_presentation:
    incremental: yes
    keep_tex: yes
    toc: yes
  ioslides_presentation:
    incremental: yes
header-includes:
- \usepackage{url}
bibliography: markovchainBiblio.bib
---

```{r setup, include=FALSE}
library(knitr)
```

<!--
%\VignetteEngine{knitr::rmarkdown }
%\VignetteIndexEntry{Crash Intro to markovchain package}
-->

#Intro

- The markovchain package [@markovchainR] will be introduced.
- The package is intended to provide S4 classes to perform probabilistic and statistical analysis of Discrete Time Markov Chains (DTMC). See [@bremaud1999discrete] for a theoretical review of the mathematics underlying the DTMC models.
- The vignette will show: how to load the package and create a DTMC, how to manage a DTMC, how to perform basic probabilistic analysis, how to fit a DTMC.

---

- The package is on Cran since Summer 2013.
- It requires a recent version of R (>=3.0). Since version 0.2 parts of code have been moved to Rcpp [@RcppR].
- The package won a slot in Google Summer of Code 2015 for optimizing internals and expanding functionalities.

---

#First moves into the markovchain package

##Loading the package
- The package is loaded using 

```{r load, echo=TRUE, warning=FALSE}
library(markovchain) #load the package
```

---

##Creating a DTMC

- DTMC can be easily create following standard S4 classes syntax. The show method displays it.

```{r create, echo=TRUE, tidy=FALSE, message=FALSE}
tmA <- matrix(c(0,0.5,0.5,.5,0,.5,.5,.5,0),nrow = 3,byrow = TRUE) #define the transition matrix
dtmcA <- new("markovchain",transitionMatrix=tmA, states=c("a","b","c"), name="MarkovChain A") #create the DTMC
dtmcA
```

- Otherwise, it can also be created directly coercing a matrix.

```{r create2, echo=TRUE}
dtmcA2<-as(tmA, "markovchain") #using coerce from matrix
states(dtmcA2) #note default names assigned to states
```

---

- It is also possible to display a DTMC, using [@pkg:igraph] capabilities

```{r plot, echo=TRUE}
plot(dtmcA)
```

---

#Probabilistic analysis

#The basic

- It is possible to access transition probabilities and to perform basic operations.
- Similarly, it is possible to access the conditional distribution of states, $Pr\left ( X_{t+1} | X_{t}=s \right )$

```{r proprieties, echo=TRUE, tidy=FALSE}
dtmcA[2,3] #using [ method
transitionProbability(dtmcA, "b","c") #using specific S4 method
conditionalDistribution(dtmcA,"b")
```

---

- It is possible to simulate states distribution after n-steps

```{r transitions, echo=FALSE, tidy=FALSE}
initialState<-c(0,1,0)
steps<-4
finalState<-initialState*dtmcA^steps #using power operator
finalState
```

---

- As well as steady states distribution

```{r steadystate, echo=FALSE, tidy=FALSE}
steadyStates(dtmcA) #S4 method
```

---
##Advanced 

- We use an example found on Mathematica Web page, [@mathematica]

```{r mathematicaMc, echo=TRUE, tidy=FALSE}
E <- matrix(0, nrow = 4, ncol = 4)
E[1, 2] <- 1;E[2, 1] <- 1/3; E[2, 3] <- 2/3
E[3,2] <- 1/4; E[3, 4] <- 3/4; E[4, 3] <- 1
mcMathematica <- new("markovchain", states = c("a", "b", "c", "d"),transitionMatrix = E,
name = "Mathematica")
```

---

- The summary method shows the proprieties of the DTCM

```{r summary, echo=TRUE, tidy=FALSE}
summary(mcMathematica)
```


#Bibliography {.allowframebreaks}
