---
title: "Crash Introduction to markovchain R package"
author: "Giorgio Alfredo Spedicato, Ph.D C.Stat ACAS"
date: "19th April 2015"
output:
  beamer_presentation:
    incremental: yes
    keep_tex: yes
    toc: yes
  ioslides_presentation:
    incremental: yes
header-includes:
- \usepackage{url}
bibliography: markovchainBiblio.bib
---

```{r setup, include=FALSE}
library(knitr)
```

<!--
%\VignetteEngine{knitr::rmarkdown }
%\VignetteIndexEntry{Crash Intro to markovchain package}
-->

#Intro

- The markovchain package [@markovchainR] will be introduced.
- The package is intended to provide S4 classes to perform probabilistic and statistical analysis of Discrete Time Markov Chains (DTMC). See [@bremaud1999discrete] for a theoretical review of the mathematics underlying the DTMC models.
- The vignette will show: how to load the package and create a DTMC, how to manage a DTMC, how to perform basic probabilistic analysis, how to fit a DTMC.

---


#First moves into the markovchain package

##Loading the package
- The package is loaded using 

```{r load, echo=TRUE, warning=FALSE, message=FALSE}
library(markovchain) #load the package
```

- It requires a recent version of R (>=3.0) and the markovchain package [@markovchainR]. The development version of the package requires also Rcpp package [@RcppR].

---

#Bibliography {.allowframebreaks}
