---
title: "A published non-rejection example for the Markov property test"
author: "Giorgio Alfredo Spedicato"
format: html
bibliography:
  - statistical_analysis_refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{A published non-rejection example for the Markov property test}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
library(markovchain)
```

## A published example: Spedicato et al. (2016)

Spedicato, Kang, Yalamanchi and Yadav present a worked example of the Markov-property test in *The markovchain Package: A Package for Easily Handling Discrete Markov Chains in R* [@spedicato2016markovchain]. The example uses the following 17-state sequence:

```{r}
#| label: published-markov-sequence

sample_sequence <- c(
  "a", "b", "a", "a", "a", "a", "b", "a", "b",
  "a", "b", "a", "a", "b", "b", "b", "a"
)

sample_sequence
```

The authors construct the contingency tables of successive triples and apply a Pearson chi-squared test. Their reported result is a chi-squared statistic of $0.28$ on 5 degrees of freedom, with p-value $0.9980024$. Thus the null hypothesis of the first-order Markov property is **not rejected** at the 5% level. The paper explicitly uses this example to illustrate how a p-value above the chosen significance level provides no evidence against the Markov property.

The same published example can be reproduced directly with the current package interface:

```{r}
#| label: published-markov-reproduction

result <- verifyMarkovProperty(
  sample_sequence,
  method = "G",
  verbose = TRUE
)

result$p.value
```

The current implementation uses the likelihood-ratio statistic by default (`method = "G"`), whereas the published example used Pearson's chi-squared formulation. The two statistics are asymptotically equivalent under the same null model, but their numerical values need not be identical. The important literature benchmark here is the published **non-rejection** of the first-order Markov hypothesis.

For comparison, the same paper also discusses the assessment of Markov order and gives examples where the null hypothesis is rejected. This makes the paper particularly useful for the vignette because it documents both sides of the inferential decision rather than presenting only rejection examples.

## Why this example replaces the Collins example

This example is preferable to the previous Collins (1974) vignette example for reproducibility: the complete numerical sequence is printed in the published source, so the reader can reconstruct the test without having to obtain undocumented micro-unit manufacturing data. It also tests exactly the first-order Markov property implemented by `verifyMarkovProperty()`, rather than relying on a qualitative statement about a published result.
