---
title: "Statistical analysis of Markov chain sequences"
author: "Giorgio Alfredo Spedicato"
format: html
bibliography:
  - markovchainBiblio.bib
  - statistical_analysis_refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Statistical analysis of Markov chain sequences}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
#| echo: false

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(123)
library(markovchain)
```

# Statistical analysis

## Testing the first-order Markov property

For an observed sequence $X_1,\ldots,X_n$ with $r$ possible states, the first-order Markov property requires that

$$
P(X_{t+1}=k\mid X_t=j,X_{t-1}=i)=P(X_{t+1}=k\mid X_t=j).
$$

Equivalently, $X_{t-1}$ and $X_{t+1}$ are conditionally independent given $X_t$. Let $N_{ijk}$ denote the number of consecutive triples $(i,j,k)$. Under the first-order null model,

$$
E_{ijk}=N_{ij+}\frac{N_{+jk}}{N_{+j+}}.
$$

This formulation is closely related to the classical likelihood-based inference for Markov chains [@anderson1957statistical; @kullback1962tests]. The package implements the Pearson statistic

$$
X^2=\sum_{i,j,k}\frac{(N_{ijk}-E_{ijk})^2}{E_{ijk}}
$$

and the likelihood-ratio statistic

$$
G^2=2\sum_{N_{ijk}>0}N_{ijk}\log\left(\frac{N_{ijk}}{E_{ijk}}\right).
$$

For complete support the asymptotic degrees of freedom are $r(r-1)^2$. With sparse support the implementation derives the degrees of freedom from the observed conditional support. A non-rejection means only that the data provide insufficient evidence against the first-order model; it does not prove the Markov property.

```{r}
#| label: markov-test

data(rain)
verifyMarkovProperty(rain$rain, method = "G", verbose = TRUE)
verifyMarkovProperty(rain$rain, method = "Pearson", verbose = TRUE)
```

The printed output above reports the statistic, degrees of freedom and p-value for the two asymptotic tests. Keeping `verbose = TRUE` in the vignette makes the numerical result part of the rendered example rather than silently suppressing it.

## Empirical versus theoretical transition probabilities

`verifyEmpiricalToTheoretical()` tests whether observed transition counts are compatible with a specified transition matrix. For each origin state $i$, the null model is multinomial with probabilities given by row $i$ of the theoretical transition matrix. The likelihood-ratio and Pearson versions use the same observed row totals and differ only in their goodness-of-fit statistic. This is the natural row-wise multinomial formulation of goodness-of-fit for transition probabilities [@kullback1962tests].

```{r}
#| label: empirical-theoretical

P <- matrix(c(.70, .20, .10,
              .20, .50, .30,
              .10, .30, .60), 3, 3, byrow = TRUE,
            dimnames = list(c("a", "b", "c"), c("a", "b", "c")))
mc <- new("markovchain", states = c("a", "b", "c"), transitionMatrix = P)
set.seed(1)
x <- rmarkovchain(1000, mc, t0 = "a")
verifyEmpiricalToTheoretical(x, mc, method = "G", verbose = TRUE)
```

The function also accepts a raw transition-count matrix:

```{r}
#| label: empirical-counts

counts <- createSequenceMatrix(x)
verifyEmpiricalToTheoretical(counts, mc, method = "Pearson", verbose = TRUE)
```

If an observed transition has positive count while the theoretical probability is zero, the theoretical model is incompatible with the data. Conversely, a zero observed count where the theoretical probability is positive is not by itself evidence against the model.

For small or sparse samples, the asymptotic chi-squared approximation can be replaced by a parametric Monte Carlo calibration. The observed row totals are retained and multinomial transition counts are simulated under the theoretical matrix:

```{r}
#| label: empirical-theoretical-simulation

verifyEmpiricalToTheoretical(
  counts, mc, method = "simulation", B = 999,
  seed = 123, verbose = TRUE
)
```

The resulting p-value uses the usual Monte Carlo correction

$$
\frac{1+B_{\geq}}{B+1}.
$$

This is a parametric Monte Carlo test, not an exact finite-sample test.

## Homogeneity across samples

`verifyHomogeneity()` tests whether several empirical sequences or transition-count matrices share the same transition probabilities. Under the null hypothesis, a common transition matrix is estimated by pooling the transition counts across samples. The test can therefore be viewed, row by row, as a multinomial homogeneity test.

```{r}
#| label: homogeneity

set.seed(123)
x1 <- rmarkovchain(500, mc, t0 = "a")
x2 <- rmarkovchain(500, mc, t0 = "b")
x3 <- rmarkovchain(500, mc, t0 = "c")
verifyHomogeneity(list(x1, x2, x3), method = "G", verbose = TRUE)
```

Pearson and likelihood-ratio versions are available:

```{r}
#| label: homogeneity-pearson

verifyHomogeneity(list(x1, x2, x3), method = "Pearson", verbose = TRUE)
```

A parametric Monte Carlo version simulates multinomial rows under the pooled transition matrix while retaining the observed number of transitions originating from each state in each sample. The pooled model is re-estimated for each replicate:

```{r}
#| label: homogeneity-simulation

verifyHomogeneity(
  list(x1, x2, x3), method = "simulation", B = 999,
  seed = 123, verbose = TRUE
)
```

These tests are useful when transition data come from different periods, regions, portfolios, or cohorts and one wants to assess whether a common homogeneous transition matrix is adequate. The likelihood-ratio formulation follows the classical contingency-table and Markov-chain framework of @kullback1962tests.

## Testing time-homogeneity of transition probabilities

`assessStationarity()` tests a different form of stability: whether the transition probabilities are constant across a specified number of consecutive blocks of the sequence. For each origin state, the transition counts in the blocks form a contingency table. The null hypothesis is that the conditional distribution of the next state, given the current state, is the same in every informative block.

This distinction is important. `verifyHomogeneity()` compares **different samples** and asks whether they share a common transition matrix, whereas `assessStationarity()` splits **one sequence into consecutive blocks** and asks whether its transition matrix is stable over time. The latter is therefore a test of time-homogeneity of a first-order Markov chain.

The test is based on the Pearson chi-squared statistic. Transition counts are retained as counts; zero counts caused merely by finite sampling are not treated as impossible transitions. Blocks in which a particular origin state is never observed contain no information about that state's conditional distribution and are therefore omitted from that state's contingency table.

```{r}
#| label: stationarity

set.seed(123)
stationary.sequence <- rmarkovchain(1000, mc, t0 = "a")
assessStationarity(stationary.sequence, nblocks = 4, verbose = TRUE)
```

The output gives the combined Pearson statistic, its degrees of freedom, and the asymptotic p-value. A large p-value does not establish stationarity; it means that the data do not provide sufficient evidence against constant transition probabilities across the blocks.

### Structural zeros

A zero transition count can have two very different meanings. A **sampling zero** is simply a transition that did not occur in the observed sample. A **structural zero** is a transition that is impossible by construction of the process. Only the latter should be excluded from the contingency table and from the corresponding degrees of freedom.

`assessStationarity()` therefore accepts the optional `structural.zeros` argument. It is a logical transition matrix with `TRUE` marking transitions that are structurally impossible. Its rows and columns correspond to the states of the sequence; when dimnames are supplied they are used to match the states explicitly.

For example, suppose state `a` can never be followed by `a`:

```{r}
#| label: stationarity-structural-zeros

structural.zeros <- matrix(FALSE, 3, 3,
                           dimnames = list(c("a", "b", "c"),
                                           c("a", "b", "c")))
structural.zeros["a", "a"] <- TRUE

structural.sequence <- c("a", "b", "a", "c", "a", "b", "c", "a",
                         "b", "c", "b", "a", "c", "b", "a", "c",
                         "a", "b", "c", "a", "b", "c", "b", "a")

assessStationarity(
  structural.sequence,
  nblocks = 4,
  structural.zeros = structural.zeros,
  verbose = TRUE
)
```

A structural-zero transition that is actually observed is rejected as inconsistent with the supplied model:

```{r}
#| label: stationarity-invalid-structural-zero

invalid.zeros <- matrix(FALSE, 3, 3,
                        dimnames = list(c("a", "b", "c"),
                                        c("a", "b", "c")))
invalid.zeros["a", "b"] <- TRUE

try(
  assessStationarity(
    structural.sequence,
    nblocks = 4,
    structural.zeros = invalid.zeros,
    verbose = TRUE
  )
)
```

The distinction between sampling and structural zeros is especially relevant for sparse transition matrices. Automatically interpreting every observed zero as a structural zero would change the null model and can substantially inflate or otherwise distort the degrees of freedom.

## Published real-data examples

The stationarity of transition probabilities has been tested on real data for decades. Anderson and Goodman (1957) provide the classical likelihood-ratio framework, while Collins (1974) applies Markov-chain inference to manufacturing establishments and tests the homogeneity of transition matrices observed over successive years [@anderson1957statistical; @collins1974micro]. Tran and Carmichael (2012) apply the likelihood-ratio approach specifically to contractor payment histories and test the assumption that transition probabilities are constant over project time [@tranCarmichael2012].

### Manufacturing establishments: Collins (1974)

Collins [@collins1974micro] applies tests of the Markov property, chain order and **homogeneity of transition matrices** to data on manufacturing establishments. The study uses micro-unit data from manufacturing establishments observed over successive years. The published analysis finds no statistically significant evidence against homogeneity for the annual transition matrices. This provides a useful real-data example in which the null hypothesis of constant transition probabilities is *not* rejected, complementing examples in which stationarity is rejected.

The original article is the appropriate source for the historical data and complete transition tables; the package vignette does not reproduce proprietary or incompletely documented micro-unit records. We therefore use the Collins result as a published benchmark rather than manufacturing a pseudo-sequence from rounded tables.

### Contractor payments: Tran and Carmichael (2012)

Tran and Carmichael [@tranCarmichael2012] provide two real construction-project case studies. Their null hypothesis is

$$
H_0:p_{jk}(\tau)=p_{jk},\qquad \tau=1,\ldots,s,
$$

against time-dependent transition probabilities. They use the Anderson--Goodman likelihood-ratio test and note that the Pearson chi-squared homogeneity test is asymptotically equivalent. Both case studies reject stationarity at the 5% level. The road project has 41 progress claims over 32 months; the steel-fabrication project has 12 claims over 12 months [@tranCarmichael2012].

The second case is particularly useful for `markovchain` because the paper publishes the empirical, month-by-month outstanding amounts. Table 4 reports the cumulative total claimed amount and amounts outstanding by more than 2, 4, 6 and 8 weeks. The values below reproduce the published table in thousands of dollars [@tranCarmichael2012].

```{r}
#| label: tran-steel-data
#| echo: true

steel <- data.frame(
  month = 1:12,
  total = c(89531, 173329, 421008, 531654, 1082009, 1358504,
            1657696, 1709917, 1801566, 1818082, 1847810, 1888942),
  over2 = c(89531, 173329, 198096, 308742, 344784, 407462,
            440712, 492933, 492933, 509449, 509449, 550581),
  over4 = c(89531, 99710, 124477, 134641, 170683, 233361,
            266611, 267075, 267075, 267075, 267075, 308207),
  over6 = c(89531, 99710, 124477, 134641, 170683, 233361,
            266611, 267075, 267075, 267075, 267075, 270101),
  over8 = c(89531, 99710, 124477, 134641, 170683, 233361,
            266611, 267075, 267075, 267075, 267075, 270101)
)

steel
```

The paper converts these outstanding amounts into transition flows $\alpha_{jk}(\tau)$ and then applies the likelihood-ratio statistic. Its reported result is

$$
-2\log(\lambda)=1636.1,
\qquad df=6\times5\times11=330,
$$

with a p-value below 0.05. The null hypothesis of time-homogeneous transition probabilities is therefore rejected [@tranCarmichael2012].

Because these are **monetary transition flows rather than individual claim-level state observations**, `assessStationarity()` should not be applied to the twelve rows as if each dollar were an independent observation. Instead, the same contingency-table calculation is represented by `verifyHomogeneity()` using the published transition-flow matrices. The following code reconstructs the six-state transition-count matrices from Table 4. The two absorbing states are included as self-transitions only to complete the transition matrices; they do not carry information about time variation.

```{r}
#| label: tran-steel-transition-counts
#| echo: true

steel.transition <- function(z) {
  m <- matrix(0, 6, 6,
              dimnames = list(as.character(0:5), as.character(0:5)))

  # States 0--3 are progressively older outstanding states;
  # states 4 and 5 are Paid and To be resolved.
  m[1, 2] <- z$over2
  m[1, 5] <- z$total - z$over2
  m[2, 3] <- z$over4
  m[2, 5] <- z$over2 - z$over4
  m[3, 4] <- z$over6
  m[3, 5] <- z$over4 - z$over6
  m[4, 5] <- z$over6 - z$over8
  m[4, 6] <- z$over8

  # Absorbing states.
  m[5, 5] <- 1
  m[6, 6] <- 1
  m
}

steel.matrices <- lapply(seq_len(nrow(steel)), function(i) {
  steel.transition(steel[i, ])
})

verifyHomogeneity(
  steel.matrices,
  method = "G",
  verbose = TRUE
)
```

The reconstructed-data calculation gives the same **qualitative conclusion** as Tran and Carmichael: the p-value is overwhelmingly small and the hypothesis of constant transition probabilities is rejected. The numerical statistic is not expected to equal 1636.1 exactly, because the paper's likelihood-ratio construction uses the monetary-flow likelihood and its own phase/time aggregation, whereas `verifyHomogeneity()` applies the package's general contingency-table formulation to the reconstructed transition-flow matrices. The comparison is therefore deliberately made at the level of the inferential conclusion rather than by asserting numerical identity.

This example also illustrates why the new `structural.zeros` facility is useful. In the owner-payment model, only transitions to the next ageing state or to an absorbing state are admissible; many cells of the $6\times6$ transition matrix are structural zeros [@tranCarmichael2012]. For a genuine individual-level sequence from the same model, these constraints can be supplied directly to `assessStationarity()` through a logical transition matrix.

```{r}
#| label: tran-steel-structural-zeros
#| echo: true

payment.structural.zeros <- matrix(TRUE, 6, 6,
  dimnames = list(as.character(0:5), as.character(0:5)))

# Allowed transitions: ageing by one state, payment, and resolution;
# states 4 and 5 are absorbing.
payment.structural.zeros[1, c(2, 5)] <- FALSE
payment.structural.zeros[2, c(3, 5)] <- FALSE
payment.structural.zeros[3, c(4, 5)] <- FALSE
payment.structural.zeros[4, c(5, 6)] <- FALSE
payment.structural.zeros[5, 5] <- FALSE
payment.structural.zeros[6, 6] <- FALSE

payment.structural.zeros
```

The two published case studies therefore provide complementary real-data evidence: Collins reports a case in which homogeneity is not rejected, whereas Tran and Carmichael report two construction-payment cases in which it is rejected. They also illustrate an important practical point: structural zeros are part of the model specification and should not be confused with transitions that merely happen not to appear in a finite sample.

## Monte Carlo calibration and validation

The three statistical tests provide asymptotic G and Pearson tests and, where appropriate, a parametric Monte Carlo alternative. Monte Carlo calibration is particularly useful when expected counts are small or the transition support is sparse.

For `verifyMarkovProperty()`, a first-order transition matrix is fitted under the null, sequences are simulated from it, and the first-order model is re-fitted in every replicate. For `verifyEmpiricalToTheoretical()` and `verifyHomogeneity()`, the simulation conditions on the observed row totals and samples multinomial transition counts under the theoretical or pooled transition probabilities. In all cases the Monte Carlo p-value is

$$
\frac{1+\sum_{b=1}^B I(T_b\ge T_{obs})}{B+1},
$$

where $T$ is the selected test statistic.

For example:

```{r}
#| label: all-simulation

verifyMarkovProperty(rain$rain, method = "simulation", B = 999,
                     seed = 123, verbose = TRUE)
verifyEmpiricalToTheoretical(counts, mc, method = "simulation", B = 999,
                             seed = 123, verbose = TRUE)
verifyHomogeneity(list(x1, x2, x3), method = "simulation", B = 999,
                  seed = 123, verbose = TRUE)
```

These procedures are parametric Monte Carlo tests, not exact finite-sample tests. A separate validation study is deliberately kept outside `R CMD check` because it is computationally expensive. It should evaluate Type-I error, power, sample-size sensitivity, state-space size, sparse transition matrices, and calibration of the asymptotic tests against the Monte Carlo alternatives. The validation script is stored under `inst/validation/` and can be run manually.

## References

::: {#refs}
:::
