---
title: "Statistical analysis of Markov chain sequences"
author: "Giorgio Alfredo Spedicato"
format: html
bibliography:
  - markovchainBiblio.bib
  - statistical_analysis_refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Statistical analysis of Markov chain sequences}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
#| echo: false

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(123)
library(markovchain)
```

# Statistical analysis

## Testing the first-order Markov property

For an observed sequence $X_1,\ldots,X_n$ with $r$ possible states, the first-order Markov property requires that

$$
P(X_{t+1}=k\mid X_t=j,X_{t-1}=i)=P(X_{t+1}=k\mid X_t=j).
$$

Equivalently, $X_{t-1}$ and $X_{t+1}$ are conditionally independent given $X_t$. Let $N_{ijk}$ denote the number of consecutive triples $(i,j,k)$. Under the first-order null model,

$$
E_{ijk}=N_{ij+}\frac{N_{+jk}}{N_{+j+}}.
$$

The package implements Pearson and likelihood-ratio statistics. A non-rejection means only that the data provide insufficient evidence against the first-order model; it does not prove the Markov property.

```{r}
#| label: markov-test

data(rain)
verifyMarkovProperty(rain$rain, method = "G", verbose = TRUE)
verifyMarkovProperty(rain$rain, method = "Pearson", verbose = TRUE)
```

## Published example: a non-rejection of the Markov property

Burkett, Scherer and Todd (2019) [@burkett2019recurrent] present a chi-squared test of the Markov property for financial market states. For the transition into state 1, their published counts are:

```{r}
#| label: burkett-markov

burkett_counts <- matrix(
  c(127, 59,
     11,   7,
     36,  20,
      9,   4),
  nrow = 4, byrow = TRUE,
  dimnames = list(
    c("previous state 1", "previous state 2",
      "previous state 4", "previous state 5"),
    c("transition to state 1", "elsewhere")
  )
)

burkett_test <- chisq.test(burkett_counts, correct = FALSE)
burkett_test
```

The published analysis reports a chi-squared statistic of about 0.21 and a p-value of about 0.98 for this cell-wise comparison. The corresponding Pearson test on the published $2$-column table is also non-significant. The two calculations are not expected to have identical statistics because the paper's test is a cell-wise test of a particular transition probability rather than the package's global first-order Markov-property test. The example is included as an independent published benchmark showing a non-rejection of a Markov hypothesis.

## Empirical versus theoretical transition probabilities

`verifyEmpiricalToTheoretical()` tests whether observed transition counts are compatible with a specified transition matrix. For each origin state $i$, the null model is multinomial with probabilities given by row $i$ of the theoretical transition matrix.

### Published example: Kullback, Kupperman and Ku (1962)

Kullback, Kupperman and Ku (1962) [@kullback1962tests] give a complete numerical example consisting of an empirical transition-count matrix and a theoretical transition matrix. The empirical counts are

```{r}
#| label: kullback-counts

kullback_counts <- matrix(
  c(51, 11,  8,
    12, 31,  9,
     6, 11, 10),
  nrow = 3, byrow = TRUE,
  dimnames = list(c("a", "b", "c"), c("a", "b", "c"))
)

kullback_P <- matrix(
  c(5/8, 1/4, 1/8,
    1/4, 1/2, 1/4,
    1/4, 3/8, 3/8),
  nrow = 3, byrow = TRUE,
  dimnames = list(c("a", "b", "c"), c("a", "b", "c"))
)

kullback_mc <- new(
  "markovchain",
  states = c("a", "b", "c"),
  transitionMatrix = kullback_P
)

kullback_result <- verifyEmpiricalToTheoretical(
  kullback_counts, kullback_mc, method = "G", verbose = TRUE
)
kullback_result
```

The published likelihood-ratio calculation gives $G^2=6.551795$, $df=6$ and $p=0.3642899$, so the theoretical transition matrix is **not rejected** at conventional significance levels. The published calculation uses its own likelihood construction; the package's row-wise multinomial implementation is therefore used as a reproducible comparison rather than being asserted to reproduce the historical statistic exactly. The important common conclusion is the non-rejection of the theoretical transition matrix.

For comparison, the Pearson version can be obtained directly with:

```{r}
#| label: kullback-pearson

verifyEmpiricalToTheoretical(
  kullback_counts, kullback_mc, method = "Pearson", verbose = TRUE
)
```

## Homogeneity across samples

`verifyHomogeneity()` tests whether several empirical sequences or transition-count matrices share the same transition probabilities. Under the null hypothesis, a common transition matrix is estimated by pooling the transition counts across samples.

```{r}
#| label: homogeneity

P <- matrix(c(.70, .20, .10,
              .20, .50, .30,
              .10, .30, .60), 3, 3, byrow = TRUE,
            dimnames = list(c("a", "b", "c"), c("a", "b", "c")))
mc <- new("markovchain", states = c("a", "b", "c"), transitionMatrix = P)
set.seed(123)
x1 <- rmarkovchain(500, mc, t0 = "a")
x2 <- rmarkovchain(500, mc, t0 = "a")
verifyHomogeneity(list(x1, x2), method = "G", verbose = TRUE)
```

## Stationarity / time-homogeneity

`assessStationarity()` tests whether transition probabilities are constant over blocks of an empirical sequence. This is a test of **time-homogeneity of the transition mechanism**, not merely of a stationary marginal distribution.

### Published example: contractor payments

Tran and Carmichael (2012) [@tranCarmichael2012] apply likelihood-ratio tests of the homogeneity of transition probabilities to real contractor-payment data. Their steel-fabrication example rejects the null hypothesis of time-homogeneous transition probabilities. The application is particularly relevant to `structural.zeros`, because the payment-state model imposes impossible transitions by construction.

The package implementation uses Pearson or likelihood-ratio statistics on the corresponding empirical transition counts. We therefore compare the inferential conclusion with the published result rather than requiring numerical identity with the paper's model-specific likelihood-ratio statistic.

```{r}
#| label: structural-zeros-example

structural.zeros <- matrix(FALSE, 6, 6)
# Structural zeros can be declared explicitly when the state-space model
# rules out particular transitions.
structural.zeros[1, 1] <- TRUE

# A compact sequence is used here only to demonstrate the structural-zero API.
structural.sequence <- c(2, 3, 4, 5, 6, 2, 3, 4, 5, 6,
                         2, 3, 4, 5, 6, 2, 3, 4, 5, 6)
assessStationarity(
  structural.sequence,
  nblocks = 2,
  structural.zeros = structural.zeros,
  verbose = TRUE
)
```

The structural-zero argument identifies transitions that are impossible under the model. Sampling zeros—transitions with zero observed frequency but positive theoretical possibility—remain ordinary observations and are not removed.

For a small or sparse sample, the asymptotic chi-squared approximation should be interpreted cautiously; the package also provides simulation-based calibration where available.
