---
title: "Statistical analysis of Markov chain sequences"
author: "Giorgio Alfredo Spedicato"
format: html
bibliography:
  - markovchainBiblio.bib
  - statistical_analysis_refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Statistical analysis of Markov chain sequences}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
#| echo: false

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(123)
library(markovchain)
```

# Statistical analysis

## Testing the first-order Markov property

For an observed sequence $X_1,\ldots,X_n$ with $r$ possible states, the first-order Markov property requires that

$$
P(X_{t+1}=k\mid X_t=j,X_{t-1}=i)=P(X_{t+1}=k\mid X_t=j).
$$

Equivalently, $X_{t-1}$ and $X_{t+1}$ are conditionally independent given $X_t$. Let $N_{ijk}$ denote the number of consecutive triples $(i,j,k)$. Under the first-order null model,

$$
E_{ijk}=N_{ij+}\frac{N_{+jk}}{N_{+j+}}.
$$

This formulation is closely related to the classical likelihood-based inference for Markov chains [@anderson1957statistical; @kullback1962tests]. The package implements the Pearson statistic

$$
X^2=\sum_{i,j,k}\frac{(N_{ijk}-E_{ijk})^2}{E_{ijk}}
$$

and the likelihood-ratio statistic

$$
G^2=2\sum_{N_{ijk}>0}N_{ijk}\log\left(\frac{N_{ijk}}{E_{ijk}}\right).
$$

For complete support the asymptotic degrees of freedom are $r(r-1)^2$. With sparse support the implementation derives the degrees of freedom from the observed conditional support. A non-rejection means only that the data provide insufficient evidence against the first-order model; it does not prove the Markov property.

```{r}
#| label: markov-test

data(rain)
verifyMarkovProperty(rain$rain, method = "G", verbose = TRUE)
verifyMarkovProperty(rain$rain, method = "Pearson", verbose = TRUE)
```

The printed output above reports the statistic, degrees of freedom and p-value for the two asymptotic tests. Keeping `verbose = TRUE` in the vignette makes the numerical result part of the rendered example rather than silently suppressing it.

## Empirical versus theoretical transition probabilities

`verifyEmpiricalToTheoretical()` tests whether observed transition counts are compatible with a specified transition matrix. For each origin state $i$, the null model is multinomial with probabilities given by row $i$ of the theoretical transition matrix. The likelihood-ratio and Pearson versions use the same observed row totals and differ only in their goodness-of-fit statistic. This is the natural row-wise multinomial formulation of goodness-of-fit for transition probabilities [@kullback1962tests].

```{r}
#| label: empirical-theoretical

P <- matrix(c(.70, .20, .10,
              .20, .50, .30,
              .10, .30, .60), 3, 3, byrow = TRUE,
            dimnames = list(c("a", "b", "c"), c("a", "b", "c")))
mc <- new("markovchain", states = c("a", "b", "c"), transitionMatrix = P)
set.seed(1)
x <- rmarkovchain(1000, mc, t0 = "a")
verifyEmpiricalToTheoretical(x, mc, method = "G", verbose = TRUE)
```

The function also accepts a raw transition-count matrix:

```{r}
#| label: empirical-counts

counts <- createSequenceMatrix(x)
verifyEmpiricalToTheoretical(counts, mc, method = "Pearson", verbose = TRUE)
```

If an observed transition has positive count while the theoretical probability is zero, the theoretical model is incompatible with the data. Conversely, a zero observed count where the theoretical probability is positive is not by itself evidence against the model.

For small or sparse samples, the asymptotic chi-squared approximation can be replaced by a parametric Monte Carlo calibration. The observed row totals are retained and multinomial transition counts are simulated under the theoretical matrix:

```{r}
#| label: empirical-theoretical-simulation

verifyEmpiricalToTheoretical(
  counts, mc, method = "simulation", B = 999,
  seed = 123, verbose = TRUE
)
```

The resulting p-value uses the usual Monte Carlo correction

$$
\frac{1+B_{\geq}}{B+1}.
$$

This is a parametric Monte Carlo test, not an exact finite-sample test.

## Homogeneity across samples

`verifyHomogeneity()` tests whether several empirical sequences or transition-count matrices share the same transition probabilities. Under the null hypothesis, a common transition matrix is estimated by pooling the transition counts across samples. The test can therefore be viewed, row by row, as a multinomial homogeneity test.

```{r}
#| label: homogeneity

set.seed(123)
x1 <- rmarkovchain(500, mc, t0 = "a")
x2 <- rmarkovchain(500, mc, t0 = "b")
x3 <- rmarkovchain(500, mc, t0 = "c")
verifyHomogeneity(list(x1, x2, x3), method = "G", verbose = TRUE)
```

Pearson and likelihood-ratio versions are available:

```{r}
#| label: homogeneity-pearson

verifyHomogeneity(list(x1, x2, x3), method = "Pearson", verbose = TRUE)
```

A parametric Monte Carlo version simulates multinomial rows under the pooled transition matrix while retaining the observed number of transitions originating from each state in each sample. The pooled model is re-estimated for each replicate:

```{r}
#| label: homogeneity-simulation

verifyHomogeneity(
  list(x1, x2, x3), method = "simulation", B = 999,
  seed = 123, verbose = TRUE
)
```

These tests are useful when transition data come from different periods, regions, portfolios, or cohorts and one wants to assess whether a common homogeneous transition matrix is adequate. The likelihood-ratio formulation follows the classical contingency-table and Markov-chain framework of @kullback1962tests.

## Testing time-homogeneity of transition probabilities

`assessStationarity()` tests a different form of stability: whether the transition probabilities are constant across a specified number of consecutive blocks of the sequence. For each origin state, the transition counts in the blocks form a contingency table. The null hypothesis is that the conditional distribution of the next state, given the current state, is the same in every informative block.

This distinction is important. `verifyHomogeneity()` compares **different samples** and asks whether they share a common transition matrix, whereas `assessStationarity()` splits **one sequence into consecutive blocks** and asks whether its transition matrix is stable over time. The latter is therefore a test of time-homogeneity of a first-order Markov chain.

The test is based on the Pearson chi-squared statistic. Transition counts are retained as counts; zero counts caused merely by finite sampling are not treated as impossible transitions. Blocks in which a particular origin state is never observed contain no information about that state's conditional distribution and are therefore omitted from that state's contingency table.

```{r}
#| label: stationarity

set.seed(123)
stationary.sequence <- rmarkovchain(1000, mc, t0 = "a")
assessStationarity(stationary.sequence, nblocks = 4, verbose = TRUE)
```

The output gives the combined Pearson statistic, its degrees of freedom, and the asymptotic p-value. A large p-value does not establish stationarity; it means that the data do not provide sufficient evidence against constant transition probabilities across the blocks.

### Structural zeros

A zero transition count can have two very different meanings. A **sampling zero** is simply a transition that did not occur in the observed sample. A **structural zero** is a transition that is impossible by construction of the process. Only the latter should be excluded from the contingency table and from the corresponding degrees of freedom.

`assessStationarity()` therefore accepts the optional `structural.zeros` argument. It is a logical transition matrix with `TRUE` marking transitions that are structurally impossible. Its rows and columns correspond to the states of the sequence; when dimnames are supplied they are used to match the states explicitly.

For example, suppose state `a` can never be followed by `a`:

```{r}
#| label: stationarity-structural-zeros

structural.zeros <- matrix(FALSE, 3, 3,
                           dimnames = list(c("a", "b", "c"),
                                           c("a", "b", "c")))
structural.zeros["a", "a"] <- TRUE

# Generate a sequence respecting the structural constraint.
structural.sequence <- c("a", "b", "a", "c", "a", "b", "c", "a",
                         "b", "c", "b", "a", "c", "b", "a", "c",
                         "a", "b", "c", "a", "b", "c", "b", "a")

assessStationarity(
  structural.sequence,
  nblocks = 4,
  structural.zeros = structural.zeros,
  verbose = TRUE
)
```

A structural-zero transition that is actually observed is rejected as inconsistent with the supplied model:

```{r}
#| label: stationarity-invalid-structural-zero

invalid.zeros <- matrix(FALSE, 3, 3,
                        dimnames = list(c("a", "b", "c"),
                                        c("a", "b", "c")))
invalid.zeros["a", "b"] <- TRUE

try(
  assessStationarity(
    structural.sequence,
    nblocks = 4,
    structural.zeros = invalid.zeros,
    verbose = TRUE
  )
)
```

The distinction between sampling and structural zeros is especially relevant for sparse transition matrices. Automatically interpreting every observed zero as a structural zero would change the null model and can substantially inflate or otherwise distort the degrees of freedom.

## Published example: Alofi daily rainfall

Avery and Henderson [@averyHenderson] analysed discrete-state time series including the Alofi daily rainfall data distributed with `markovchain`. The data classify successive days into three rainfall states and are included as `rain`.

```{r}
#| label: alofi-data

data(rain)
table(rain$rain)
createSequenceMatrix(rain$rain)
```

A published likelihood comparison of first- and second-order Markov models for these data reports maximized log-likelihoods of approximately -1038.06 and -1025.10, respectively [@davison2003statistical]. Thus

$$
G^2=2\{(-1025.10)-(-1038.06)\}=25.92.
$$

For three states the comparison has 12 degrees of freedom, giving an asymptotic p-value of approximately 0.011. Thus the first-order model is rejected at the 5% level in that likelihood comparison.

The package performs the corresponding first-order goodness-of-fit calculation directly from the sequence:

```{r}
#| label: alofi-test

verifyMarkovProperty(rain$rain, method = "G", verbose = TRUE)
```

The exact value need not equal the rounded published likelihood calculation, but the hypothesis is the same: whether the state immediately preceding the current rainfall state contains additional information about the next state.

## Monte Carlo calibration and validation

The three statistical tests provide asymptotic G and Pearson tests and, where appropriate, a parametric Monte Carlo alternative. Monte Carlo calibration is particularly useful when expected counts are small or the transition support is sparse.

For `verifyMarkovProperty()`, a first-order transition matrix is fitted under the null, sequences are simulated from it, and the first-order model is re-fitted in every replicate. For `verifyEmpiricalToTheoretical()` and `verifyHomogeneity()`, the simulation conditions on the observed row totals and samples multinomial transition counts under the theoretical or pooled transition probabilities. In all cases the Monte Carlo p-value is

$$
\frac{1+\sum_{b=1}^B I(T_b\ge T_{obs})}{B+1},
$$

where $T$ is the selected test statistic.

For example:

```{r}
#| label: all-simulation

verifyMarkovProperty(rain$rain, method = "simulation", B = 999,
                     seed = 123, verbose = TRUE)
verifyEmpiricalToTheoretical(counts, mc, method = "simulation", B = 999,
                             seed = 123, verbose = TRUE)
verifyHomogeneity(list(x1, x2, x3), method = "simulation", B = 999,
                  seed = 123, verbose = TRUE)
```

These procedures are parametric Monte Carlo tests, not exact finite-sample tests. A separate validation study is deliberately kept outside `R CMD check` because it is computationally expensive. It should evaluate Type-I error, power, sample-size sensitivity, state-space size, sparse transition matrices, and calibration of the asymptotic tests against the Monte Carlo alternatives. The validation script is stored under `inst/validation/` and can be run manually.

## References

::: {#refs}
:::
