---
title: "Statistical analysis of Markov chain sequences"
author: "Giorgio Alfredo Spedicato"
format: html
bibliography: statistical_analysis_refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Statistical analysis of Markov chain sequences}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
#| echo: false

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(123)
library(markovchain)
```

# Statistical analysis

This vignette illustrates the statistical inference functions for empirical
Markov chains. Test results are returned as objects of class `htest`, following
R's standard convention for hypothesis tests.

## Testing the first-order Markov property

For an observed sequence $X_1,\ldots,X_n$, the first-order Markov property
requires that

$$
P(X_{t+1}=k\mid X_t=j,X_{t-1}=i)=P(X_{t+1}=k\mid X_t=j).
$$

The package implements Pearson and likelihood-ratio tests of this property.
A non-rejection means that the data provide insufficient evidence against the
null hypothesis; it does not prove that the process is Markovian.

```{r}
#| label: markov-test

sequence <- c("a", "b", "a", "a", "a", "a", "b", "a", "b",
              "a", "b", "a", "a", "b", "b", "b", "a")

markov_test <- verifyMarkovProperty(sequence, method = "G", verbose = TRUE)
markov_test
class(markov_test)
```

## Published example: non-rejection of the Markov property

Burkett, Scherer and Todd (2019) [@burkett2019recurrent] study recurrent and
persistent financial-market states. One of their published tests considers
the transition into state 1 and reports the following counts:

```{r}
#| label: burkett-markov

burkett_counts <- matrix(
  c(127, 59,
     11,  7,
     36, 20,
      9,  4),
  nrow = 4,
  byrow = TRUE,
  dimnames = list(
    c("previous state 1", "previous state 2",
      "previous state 4", "previous state 5"),
    c("transition to state 1", "transition elsewhere")
  )
)

chisq.test(burkett_counts, correct = FALSE)
```

The published contingency table gives a non-significant Pearson test
($\chi^2\approx0.64$, 3 degrees of freedom, p-value about 0.89). It is
included as an independent literature example of a non-rejection; it is not
claimed to be numerically identical to `verifyMarkovProperty()`, because the
published test is a cell-wise test for a particular transition rather than the
package's global first-order Markov-property test.

## Empirical versus theoretical transition probabilities

`verifyEmpiricalToTheoretical()` tests whether observed transition counts are
compatible with a specified transition matrix. The following example is from
Kullback, Kupperman and Ku (1962) [@kullback1962tests]. The paper gives both
the empirical counts and the theoretical transition matrix.

```{r}
#| label: kullback-example

kullback_counts <- matrix(
  c(51, 11,  8,
    12, 31,  9,
     6, 11, 10),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(c("a", "b", "c"), c("a", "b", "c"))
)

kullback_P <- matrix(
  c(5/8, 1/4, 1/8,
    1/4, 1/2, 1/4,
    1/4, 3/8, 3/8),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(c("a", "b", "c"), c("a", "b", "c"))
)

kullback_mc <- new(
  "markovchain",
  states = c("a", "b", "c"),
  transitionMatrix = kullback_P
)

kullback_result <- verifyEmpiricalToTheoretical(
  kullback_counts,
  kullback_mc,
  method = "G",
  verbose = TRUE
)

kullback_result
class(kullback_result)
```

Kullback et al. report $G^2=6.551795$, $df=6$ and $p=0.3642899$.
The historical `markovchain` implementation reproduced the published value
by using column totals when constructing expected counts. For the row-wise
multinomial null implemented by the current package, expected counts are
constructed as

$$E_{ij}=n_{i+}P_{ij},$$

where $n_{i+}$ is the observed total in row $i$. This gives
$G^2=6.518384$, with $df=6$ and the same non-rejection of the null hypothesis.
The small numerical discrepancy is therefore a correction of the historical
implementation rather than a change in the inferential conclusion.

## Homogeneity across samples

`verifyHomogeneity()` tests whether several empirical transition-count
matrices are compatible with a common transition mechanism.

```{r}
#| label: homogeneity

P <- matrix(
  c(.70, .20, .10,
    .20, .50, .30,
    .10, .30, .60),
  nrow = 3,
  byrow = TRUE,
  dimnames = list(c("a", "b", "c"), c("a", "b", "c"))
)

mc <- new("markovchain",
          states = c("a", "b", "c"),
          transitionMatrix = P)

set.seed(123)
x1 <- rmarkovchain(500, mc, t0 = "a")
x2 <- rmarkovchain(500, mc, t0 = "a")

homogeneity_test <- verifyHomogeneity(
  list(x1, x2), method = "G", verbose = TRUE
)
homogeneity_test
```

## Stationarity and time-homogeneity

`assessStationarity()` tests whether transition probabilities remain constant
across blocks of an empirical sequence. This is a test of time-homogeneity of
the transition mechanism, rather than a test of the marginal stationary
distribution.

### Published example: contractor payments

Tran and Carmichael (2012) [@tranCarmichael2012] apply likelihood-ratio tests
to transition probabilities in a real contractor-payment application. Their
analysis provides an example where the null hypothesis of time-homogeneous
transition probabilities is rejected. The application also motivates explicit
handling of structural zeros, i.e. transitions that are impossible under the
state-space model.

The package allows such transitions to be specified through the
`structural.zeros` argument.

```{r}
#| label: structural-zeros

structural.zeros <- matrix(FALSE, 3, 3,
                           dimnames = list(c("a", "b", "c"),
                                           c("a", "b", "c")))
structural.zeros["a", "a"] <- TRUE

structural_sequence <- c(
  "a", "b", "a", "c", "a", "b", "c", "a",
  "b", "c", "b", "a", "c", "b", "a", "c",
  "a", "b", "c", "a", "b", "c", "b", "a"
)

stationarity_test <- assessStationarity(
  structural_sequence,
  nblocks = 4,
  structural.zeros = structural.zeros,
  verbose = TRUE
)
stationarity_test
class(stationarity_test)
```

A structural zero represents a transition ruled out by the model. It is
therefore different from a sampling zero: an unobserved but possible
transition remains part of the statistical analysis.

For sparse data, the asymptotic chi-squared approximation should be interpreted
with care.
