---
title: "Statistical analysis of Markov chain sequences"
author: "Giorgio Alfredo Spedicato"
output: rmarkdown::html_vignette
vignette: >
  %\\VignetteIndexEntry{Statistical analysis of Markov chain sequences}
  %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}
bibliography: markovchainBiblio.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(123)
library(markovchain)
```

# Statistical analysis

## Testing the first-order Markov property

For an observed sequence $X_1,\\ldots,X_n$ with $r$ possible states, the first-order Markov property requires that

$$P(X_{t+1}=k\\mid X_t=j,X_{t-1}=i)=P(X_{t+1}=k\\mid X_t=j).$$

Equivalently, $X_{t-1}$ and $X_{t+1}$ are conditionally independent given $X_t$. Let $N_{ijk}$ denote the number of consecutive triples $(i,j,k)$. Under the first-order null model,

$$E_{ijk}=N_{ij+}\\frac{N_{+jk}}{N_{+j+}}.$$

The package implements the Pearson statistic

$$X^2=\\sum_{i,j,k}\\frac{(N_{ijk}-E_{ijk})^2}{E_{ijk}}$$

and the likelihood-ratio statistic

$$G^2=2\\sum_{N_{ijk}>0}N_{ijk}\\log\\left(\\frac{N_{ijk}}{E_{ijk}}\\right).$$

For complete support the asymptotic degrees of freedom are $r(r-1)^2$. With sparse support the implementation derives the degrees of freedom from the observed conditional support. A non-rejection means only that the data provide insufficient evidence against the first-order model; it does not prove the Markov property.

```{r markov-test}
data(rain)
verifyMarkovProperty(rain$rain, method = "G", verbose = FALSE)
verifyMarkovProperty(rain$rain, method = "Pearson", verbose = FALSE)
```

## Empirical versus theoretical transition probabilities

`verifyEmpiricalToTheoretical()` tests whether observed transition counts are compatible with a specified transition matrix. For each origin state $i$, the null model is multinomial with probabilities given by row $i$ of the theoretical transition matrix.

```{r empirical-theoretical}
P <- matrix(c(.70, .20, .10,
              .20, .50, .30,
              .10, .30, .60), 3, 3, byrow = TRUE,
            dimnames = list(c("a", "b", "c"), c("a", "b", "c")))
mc <- new("markovchain", states = c("a", "b", "c"), transitionMatrix = P)
set.seed(1)
x <- rmarkovchain(1000, mc, t0 = "a")
verifyEmpiricalToTheoretical(x, mc, method = "G", verbose = FALSE)
```

The function also accepts a raw transition-count matrix:

```{r empirical-counts}
counts <- createSequenceMatrix(x)
verifyEmpiricalToTheoretical(counts, mc, method = "Pearson", verbose = FALSE)
```

The likelihood-ratio formulation is the standard contingency-table approach to comparing observed multinomial counts with a specified probability vector. If an observed transition has positive count while the theoretical probability is zero, the theoretical model is incompatible with the data.

## Homogeneity across samples

`verifyHomogeneity()` tests whether several empirical sequences share the same transition probabilities. For each origin state, the individual rows form a multinomial homogeneity problem and the pooled row provides the null transition probabilities.

```{r homogeneity}
set.seed(123)
x1 <- rmarkovchain(500, mc, t0 = "a")
x2 <- rmarkovchain(500, mc, t0 = "b")
x3 <- rmarkovchain(500, mc, t0 = "c")
verifyHomogeneity(list(x1, x2, x3), method = "G", verbose = FALSE)
```

This test is useful when transition data come from different periods, regions, portfolios, or cohorts and one wants to assess whether a common homogeneous transition matrix is adequate. The likelihood-ratio formulation follows the classical contingency-table and Markov-chain framework of Kullback, Kupperman and Ku (1962).

## Published example: Alofi daily rainfall

Avery and Henderson (1999) analysed discrete-state time series including the Alofi daily rainfall data distributed with `markovchain`. The data classify successive days into three rainfall states and are included as `rain`.

```{r alofi-data}
data(rain)
table(rain$rain)
createSequenceMatrix(rain$rain)
```

The same data are discussed by Davison (2003) when comparing Markov models of different orders. His reported maximized log-likelihoods for first- and second-order models are approximately -1038.06 and -1025.10, giving

$$G^2=2\\{(-1025.10)-(-1038.06)\\}=25.92.$$

For three states the comparison has 12 degrees of freedom, giving an asymptotic p-value of approximately 0.011. Thus the first-order model is rejected at the 5% level in that published likelihood comparison.

The package performs the corresponding first-order goodness-of-fit calculation directly from the sequence:

```{r alofi-test}
verifyMarkovProperty(rain$rain, method = "G", verbose = FALSE)
```

The exact value need not equal the rounded published likelihood calculation, but the hypothesis is the same: whether the state immediately preceding the current rainfall state contains additional information about the next state.

## Monte Carlo calibration and validation

For sparse sequences, the asymptotic chi-squared approximation can be poor. `verifyMarkovProperty()` therefore provides a parametric Monte Carlo option:

```{r alofi-simulation}
verifyMarkovProperty(rain$rain, method = "simulation", B = 999,
                     seed = 123, verbose = FALSE)$p.value
```

Under the null, a first-order transition matrix is fitted, sequences are simulated from it, and the first-order model is re-fitted in every replicate. The Monte Carlo p-value is

$$\\frac{1+\\sum_{b=1}^B I(G_b^2\\ge G_{obs}^2)}{B+1}.$$

This is a parametric Monte Carlo test, not an exact finite-sample test.

A larger Monte Carlo validation study is deliberately kept outside `R CMD check` because it is computationally expensive. It should evaluate at least:

1. **Type-I error:** simulate under the null and estimate rejection rates at 1%, 5% and 10%.
2. **Power:** simulate second-order alternatives with controlled departure from first-order behaviour.
3. **Sample-size sensitivity:** repeat for several sequence lengths, e.g. 50, 100, 250, 500 and 1000.
4. **State-space sensitivity:** use 2, 3 and 5 states.
5. **Sparsity:** include transition matrices with structural zeros and highly unbalanced probabilities.
6. **Calibration:** compare asymptotic Pearson/G tests with the parametric Monte Carlo p-value.

The validation script is stored under `inst/validation/` and should be run manually rather than as a package test.

## References

Anderson, T. W. and Goodman, L. A. (1957). Statistical inference about Markov chains. *The Annals of Mathematical Statistics*, 28(1), 89--110.

Avery, P. J. and Henderson, D. A. (1999). Fitting Markov chain models to discrete state series. *Applied Statistics*, 48, 53--61.

Davison, A. C. (2003). *Statistical Models*. Cambridge University Press, Section 6.1.

Kullback, S., Kupperman, M. and Ku, H. H. (1962). Tests for contingency tables and Markov chains. *Technometrics*, 4(4), 573--608.
